### 2011-11-27 AffineLieAlgebras/Thesis/5.tex
Thesis partially structure cleaned, some translations in chapter 2.
Chapter 2 is completely translated. Description of computational package is moved to the last chapter.

\chapter{Конформная теория поля}
\label{cha:cft}

Подробное изложение \cite{difrancesco1997cft}.

\section{Общие свойства}
\label{sec:cft-general}
Алгебра Вирасоро. Представления алгебры Вирасоро. Модулярная инвариантность.

\section{Модели Весса-Зумино-Новикова-Виттена}
\label{sec:wzw}
Связь с аффинными алгебрами Ли.
Конформные вложения и недиагональные модулярные инварианты.

%\usepackage{verbatim} 
\newenvironment{comment}
{\par\noindent{\bf Для заинтересованных читателей}\\}
{\\\setlength{10cm}\\\hfill$\scriptstyle\blacksquare$\par}

\begin{abstract}
Примерный текст доклада на семинаре кафедры ФВЭиЭЧ по работе ``Recursive algorithms, branching coefficients and applications''.
\end{abstract}

\section{Постановка задачи}
\label{sec:task}

Здравствуйте!
Сегодня я хочу рассказать о работе ``Рекуррентные алгоритмы, коэффициенты ветвления аффинных алгебр
Ли и приложения''. В этой работе мы решали следующую задачу.

Даны две аффинные алгебры Ли $\mathfrak{a}\subset \mathfrak{g}$. Аффинные алгебры Ли --- это особый
класс бесконечномерных алгебр, которые строятся из конечномерных (полу)простых алгебр Ли, как
центральные расширения алгебр петель. Более подробное определение будет дано позже. Неприводимое
представление $L^{(\mu)}_{\mathfrak{g}}$ аффинной алгебры $\mathfrak{g}$ со старшим весом $\mu$
может быть разложено на представления подалгебры $\mathfrak{a}$. Наша задача --- придумать
практический алгоритм вычисления коэффициентов этого разложения. Конечно же эта задача уже решалась,
например, существует способ решения ``в лоб'', требующий построения всех представлений и различные способы,
которые работают для частных случаев, например, многое упрощается в физически интересном случае
конформных вложений. Мы предлагаем общий метод, но иллюстрировать его будем как-раз физическими
примерами конформных вложений.

Я начну с физической мотивации этой задачи в двумерной конформной теории поля, напомню некоторые
факты из теории представлений аффинных алгебр Ли, кратко опишу предложенное нами решение проблемы и
расскажу про физические примеры в нашей статье.

\section{Физическая мотивация}
\label{sec:physics}

Теперь я перейду к краткому объяснению метода и примерам вычислений.

\section{Рекуррентные соотношения для коэффициентов ветвления}
\label{sec:branching}

Ранее для вычисления коэффициентов ветвления был предложен способ, основанный на преобразовании
формулы Вейля-Каца для характеров представлений аффинных алгебр Ли в рекуррентное соотношение для
коэффициентов ветвления. Однако исходный алгоритм не работал для не максимальных вложений. В данной
работе была решена данная проблема и предложено обобщение рекуррентных соотношений.

Используя формулу Вейля-Каца мы выделяем два набора элементов в весовом пространстве. Первый набор $\Gamma$
называется веером и описывает вложение подалгебры $\mathfrak{a}$ в алгебру $\mathfrak{g}$. Веер
универсален, в том смысле, что он может быть использован для разложения любого из представлений.
Другой набор элементов весового пространства $\Psi ^{\left( \mu \right) }$ называется набором
аномальных весов. Он описывает раскладываемое представление без явного его построения.

После этого процесс вычисления коэффициентов ветвления можно реализовать при помощи следующей
рекуррентной формулы:
\begin{equation}
  \label{eq:1}
  \begin{array}{c}
      k_{\xi }^{\left( \mu \right) }=-\frac{1}{s\left( \gamma _{0}\right) }\left(
  \sum_{\omega\in W_{\bot}\backslash W} \epsilon(\omega)\; {\rm dim}\left(L^{\pi_{\mathfrak{a}_{\bot}}(\omega(\mu+\rho))-\rho_{\mathfrak{a}_{\bot}}}_{\mathfrak{a}_{\bot}}\right) \delta_{\xi-\gamma_0,\pi_{\mathfrak{a}}(\omega(\mu+\rho)-\rho)}+ \right.\\
\left.
+\sum_{\gamma \in
\Gamma _{\frak{a}\subset \frak{g}}}s\left( \gamma +\gamma _{0}\right) k_{\xi
+\gamma }^{\left( \mu \right) }\right)

  \end{array}
\end{equation}
Здесь $\mathfrak{a}_{\bot}$ --- это подалгебра, ``натянутая'' на корни $\mathfrak{g}$, ортогональные
к корневой системе подалгебры  $\mathfrak{a}$, $W_{\bot}$ --- соответствующая этим корням группа Вейля,
$\Gamma_{\mathfrak{a}\subset \mathfrak{g}}$ --- веер --- набор весов в разложении знаменателя $$\prod_{\alpha\in \Delta^{+}\setminus \Delta^{+}_{\mathfrak{a}_{\bot}}}
(1-e^{-\pi_{\mathfrak{a}}(\alpha)})^{\mathrm{mult}(\alpha)-\mathrm{mult}_{\mathfrak{a}}(\pi_{\mathfrak{a}}(\alpha))}$$, а $s(\gamma)$ ---
коэффициент при  $e^{\gamma}$ в этом разложении.

Набор аномальных весов $\Psi^{(\mu)}$ дает член с $\delta$-символом.

Это рекуррентное соотношение позволяет нам сформулировать следующий алгоритм вычисления
коэффициентов ветвления, который осуществляется без явного построения исходного представления или
представлений, входящих в разложение.
\begin{enumerate}
\item Построить наборы положительных корней $\Delta^{+}$ и $\Delta_{\mathfrak{a}}^{+}$.
\item Выбрать те положительные корни $\alpha\in \Delta^{+}$, которые ортогональны корневому
  подпространству  $\mathfrak{a}$ и собрать их в набор $\Delta^{+}_{\bot}$.
\item Построить множество весов $\Gamma$, которое мы называем ``веером''.
\item Найти множество $\widehat{\Psi^{(\mu)}}=\left\{\omega(\mu+\rho)-\rho;\; \omega\in W\right\}$
  аномальных весов представления $L^{(\mu)}_{\mathfrak{g}}$.
\item Выбрать в нем веса $\left\{ \lambda=\omega(\mu+\rho) | \pi_{\mathfrak{a}_{\bot}}\lambda \in
    \bar{C}_{\mathfrak{a}_{\bot}} \right\}$. Это можно легко сделать, так как у нас есть набор
  $\Delta^{+}_{\bot}$ и мы можем проверить, лежит ли вес $\pi_{\mathfrak{a}_{\bot}}\lambda$ в
  главной камере Вейля алгебры $\mathfrak{a}_{\bot}$. Для этого все скалярные произведения этого
  веса $\lambda$ с корнями из  $\Delta^{+}_{\bot}$ должны быть неотрицательны.
\item Далее для весов $\lambda=\omega(\mu+\rho),\; \pi_{\mathfrak{a}_{\bot}}\lambda\in
  \bar{C}_{\mathfrak{a}_{\bot}}$ мы вычисляем размерность соответствующих представлений
  $\mathrm{dim}\left(L^{\pi_{\mathfrak{a}_{\bot}}(\omega(\mu+\rho))-\rho_{\mathfrak{a}_{\bot}}}_{\mathfrak{a}_{\bot}}\right)$
  при помощи формулы Вейля и набора $\Delta^{+}_{\bot}$.
\item Далее рекуррентно вычисляются аномальные коэффициенты ветвления, а внутри главной камеры Вейля
  алгебры $\mathfrak{a}$ аномальные коэффициенты совпадают с настоящими коэффициентами ветвления.
\end{enumerate}

Сейчас я поясню всю процедуру на примере редукции представления простой алгебры $B_2$ ($so(5)$) на
подалгебру $A_1$ ($so(3)$), а потом расскажу про аффинные алгебры Ли.

\subsection{Примеры}
\label{sec:examples}

Я начну с примера вычисления в случае конечномерных простых алгебр Ли, так как его легко нарисовать.

Рассмотрим регулярное вложение  $A_1$ в $B_2$. Вот на рисунке корневая системы алгебры $B_2=so(5)$.
Корень  $\beta$ вложенной алгебры $A_1$ равен $\alpha_1+2\alpha_2$.

Функция $s(\gamma)$ и веер $\Gamma$ строятся исходя из корневых систем $\mathfrak{g}, \mathfrak{a}$
$\frac{\pi_{\mathfrak{a}}\left(\prod_{\alpha\in \Delta^{+}_{\mathfrak{g}}\setminus
      \Delta^{+}_{\mathfrak{a}}} \left(1-e^{-\alpha}\right) \right)}{\prod_{\beta\in
    \Delta^{+}_{\mathfrak{a}}} \left(1-e^{-\beta}\right) }$ и получается
\begin{equation}
  \label{eq:22}
  (1,2),\; (2,-1)
\end{equation}
Вторая компонента - значение $s(\gamma)$.

Опишем редукцию первого фундаментального представления $B_2$. Старший вес изображен на рисунке.
\begin{figure}[ph]
  \noindent\centering{
    \includegraphics[width=90mm]{B2_A1}
  }
  \caption{Регулярное вложение $A_1$ в $B_2$}
  \label{fig:B2_A1}
\end{figure}
Вот набор аномальных весов $\omega(\mu+\rho)-\rho,\; \omega\in W$ с соответствующими четностями
отражений $\epsilon(\omega)$.
Теперь мы факторизуем группу Вейля $W$ по подгруппе $W_{\bot}=\left\{\omega_1\right\}$. Получаем
следующий набор весов и размерностей представлений  $L^{\pi_{\mathfrak{a}_{\bot}}(\omega(\mu+\rho))-\rho_{\mathfrak{a}_{\bot}}}_{\mathfrak{a}_{\bot}}$:
\begin{figure}[h!tb]
  \noindent\centering{
    \includegraphics[width=90mm]{B2_A1_2}
  }
  \caption{Аномальные веса и соответствующие $\mathfrak{a}_{\bot}=A_1$-модули}
  \label{fig:B2_A1_2}
\end{figure}

Затем мы проектируем эти веса и размерности на корневое подпространство подалгебры
$\mathfrak{a}=A_1$ и получаем следующие аномальные веса и их кратности:
\begin{equation}
  \label{eq:25}
  (1,2),\; (0,-3),\; (-4,3),\; (-5,-2)
\end{equation}

Аномальный коэффициент ветвления  $k^{(1,0)}_{1}=2$, для коэффициента $k^{(1,0)}_{0}$ из
рекуррентного соотношения получаем
\begin{equation}
  \label{eq:23}
  k^{(1,0)}_{0}=-1\cdot k^{(1,0)}_2 +2\cdot k^{(1,0)}_1 - 3\cdot \delta_{0,0} = 1
\end{equation}

В работе рассмотрено большое число примеров вложений аффинных алгебр Ли. На самом деле мы написали
программу, которая может строить такие примеры автоматически для алгебр Ли не слишком большого
ранга.

Я сейчас покажу результаты применения нашего алгоритма для одного из примеров конформных вложений.

Рассмотрим такое вложение $\mathfrak{a}\subset \mathfrak{g}$, где
$\mathfrak{a}=\hat{A_1}\oplus \hat{A_1}$, а $\mathfrak{g}= \hat{A_3}$. Это
вложение является аффинным расширением специального вложения конечномерных алгебр Ли $A_1\oplus
A_1\subset A3$. ($su(2)\oplus su(2)\subset su(4)$).
\begin{comment}
 Это специальное вложение строится следующим образом: берем четырехмерное
представление $A_1\oplus A_1$ со старшим весом $(1,1)$. Вот веса этого представления
\ref{fig:A_1+A_1_to_A3}, их координаты в базисе фундаментальных весов: $\nu_1=(1,1),\; \nu_2=(-1,1),\; \nu_3=(1,-1),\; \nu_4=(-1,-1)$.

\begin{figure}[h]
  \begin{multicols}{2}
    \hfill
    \includegraphics[width=50mm]{A_1+A_1_to_A3}
    \hfill
    \caption{Представление для специального вложения $A_1\oplus A_1\subset A3$}
    \label{fig:A_1+A_1_to_A3}
    \hfill
    \includegraphics[width=50mm]{A1+A1-A3}
    \hfill
    \caption{Вложенные корни $A_1\oplus A_1\subset A3$}
    \label{fig:A1+A1-A3}
  \end{multicols}
\end{figure}

Тогда матричные элементы представления генераторов подалгебры Картана $b_1,b_2$ в базисе Вейля
даются выражениями
\begin{equation}
  d(b_i)=\mathrm{diag}\left(\frac{2(\nu_1,\alpha_i)}{(\alpha_i,\alpha_i)},\frac{2(\nu_2,\alpha_i)}{(\alpha_i,\alpha_i)},\frac{2(\nu_3,\alpha_i)}{(\alpha_i,\alpha_i)},\frac{2(\nu_4,\alpha_i)}{(\alpha_i,\alpha_i)}\right),
\end{equation}
так что $d(b_1)=\mathrm{diag}(1,-1,1,-1),\;
  d(b_2)=\mathrm{diag}(1,1,-1,-1).$
\end{comment}
Вложенные корни $\alpha_1,\alpha_2$ подалгебры $A_1\oplus A_1$
выражаются через корни $\tilde{\alpha}_i$ алгебры  $A_3$
\begin{equation}
  \label{eq:37}
  \begin{array}{l}
     \alpha_1=\frac{1}{2}(\tilde{\alpha}_1+\tilde{\alpha}_3)\\
     \alpha_2=\frac{1}{2}(\tilde{\alpha}_1+2\tilde{\alpha}_2+\tilde{\alpha}_3)
  \end{array}
\end{equation}
\begin{comment}
Они изображены на рисунке \ref{fig:A1+A1-A3}.
\end{comment}

Вложение имеет индексы  $(2,2)$ и является конформным, так как $c(A_1\oplus A_1)=c(A_1)+c(A_1)=2\frac{x_e \mathrm{dim}(A_1)}{x_e+2}=\frac{\mathrm{dim}A_3}{5}=c(A_3)$.

Для построения модулярно-инвариантной статсуммы нам нужно знать редукцию фундаментальных
представлений $\widehat{su(4)}$. Фундаментальные веса имеют следующие координаты в ортогональном базисе
\begin{equation}
  \begin{array}{lll}
     \omega_0 & = & (0,0,0,0;1;0)\\
     \omega_1 & = & (\frac{3}{4},-\frac{1}{4},-\frac{1}{4},-\frac{1}{4}; 1; 0)\\
     \omega_2 & = & (\frac{1}{2},\frac{1}{2},-\frac{1}{2},-\frac{1}{2}; 1; 0)\\
     \omega_3 & = & (\frac{1}{4},\frac{1}{4},\frac{1}{4},-\frac{3}{4}; 1; 0) \\
  \end{array}
\end{equation}

Множества положительных корней  $\Delta^{+}$ и $\Delta_{\mathfrak{a}}^{+}$:
\begin{equation}
  \label{eq:38}
  \begin{array}{lll}
    \Delta^{+} &=&\left\{\co{\Delta}^{+}=\{\tilde{\alpha}_1, \tilde{\alpha}_2, \tilde{\alpha}_3, \tilde{\alpha}_1+\tilde{\alpha}_2, \tilde{\alpha}_2+\tilde{\alpha}_3, \tilde{\alpha}_1+\tilde{\alpha}_2+\tilde{\alpha}_3\};\right.\\
    &\; & \quad\left.\co{\Delta}+n\delta;\;     +n\delta\;\mbox{с кратностью}\; 3;\; n=1,2,\dots,\right\}\\
    \Delta_{\mathfrak{a}}^{+} &=& \{  \alpha_1,\alpha_2;\;\pm \alpha_1-n\delta,\pm
    \alpha_2-n\delta;\; -n\delta\; \mbox{с кратностью} \; 2;\; n=1,2,\dots\}\\
  \end{array}
\end{equation}

Набор $\Delta^{+}_{\bot}$ пуст.
Веер  $\Gamma _{\frak{a}\subset \frak{g}}$ показан на рисунке \ref{fig:A1+A1-A3_fan}. Конечные координаты
даны в базисе фундаментальных весов $A_1 \oplus A_1$. Элемент $\gamma$ обозначен крестом, если
$s(\gamma)=1$ и кругом, если $s(\gamma)=-1$.
\begin{figure}[h!tb]
  \includegraphics[width=150mm]{A1+A1-A3_fan}
  \caption{Веер для специального вложения $\hat A_1\oplus\hat A_1\subset\hat A_3$}
  \label{fig:A1+A1-A3_fan}
\end{figure}

Мы ограничимся вычислениями до пятого грейда.
\begin{figure}[h!tb]
  \includegraphics[width=170mm]{A1+A1-A3_anom}
  \caption{Спроектированные аномальные веса $L^{\omega_2}_{A_3}$}
  \label{fig:A1+A1-A3_anom}
\end{figure}

Множество аномальных весов  $\widehat{\Psi^{(\mu)}}=\left\{\omega(\mu+\rho)-\rho;\; \omega\in
  W\right\}$ представления алгебры $\hat{A}_3$ состоит из 192 элементов, их проекции
$\pi_{\mathfrak{a}}\left(\widehat{\Psi^{(\mu)}}\right)$ показаны на рисунке  \ref{fig:A1+A1-A3_anom}
для $\mu=\omega_2=(0,1,0;1;0)$.

Аналогичные картинки для $\omega_0, \omega_1,\omega_3$ я не показываю.

Аномальные коэффициенты ветвления для представления  $L^{(\omega_2)}$ вычислены при помощи
рекуррентных соотношений и показаны на рисунке \ref{fig:A1+A1-A3_branching}. 
\begin{figure}[h!tb]
  \hspace*{-2cm}\includegraphics[width=180mm]{A1+A1-A3_branching}
  \caption{Аномальные коэффициенты ветвления для представления $L^{(0,1,0;1;0)}_{A_3}$}
  \label{fig:A1+A1-A3_branching}
\end{figure}

>>>>

\section{Вычислительные методы для теории представлений аффинных алгебр Ли}
\label{cha:computational-methods}
%% \newcommand{\go}{\stackrel{\circ }{\mathfrak{g}}}
%% \newcommand{\ao}{\stackrel{\circ }{\mathfrak{a}}}
%% \newcommand{\co}[1]{\stackrel{\circ }{#1}}
%% \newcommand{\pia}{\pi_{\mathfrak{a}}}
%% \newcommand{\piab}{\pi_{\mathfrak{a}_{\bot}}}
%% \newcommand{\gf}{\mathfrak{g}}
%% \newcommand{\af}{\mathfrak{a}}
%\newcommand{\bff}{\mathfrak{b}}
%% \newcommand{\afb}{\mathfrak{a}_{\bot}}
%% \newcommand{\hf}{\mathfrak{h}}
%\newcommand{\hfg}{\hf_{\gf}}
%% \newcommand{\hfb}{\mathfrak{h}_{\bot}}
%% \newcommand{\pf}{\mathfrak{p}}
%% \newcommand{\aft}{\widetilde{\mathfrak{a}}}

\begin{abstract}
%% Text of abstract
In this paper we present {\bf Affine.m} -- program for computations in representation theory of finite-dimensional and affine Lie algebras and describe implemented algorithms.  Algorithms are based on the properties of weights and Weyl symmetry. The most important problems for us are the ones, concerning computation of weight multiplicities in irreducible and Verma modules, branching of representations and tensor product decomposition. These problems have numerous applications in physics and we provide some examples of these applications. The program is implemented in popular computer algebra system {\it Mathematica} and works with finite-dimensional and affine Lie algebras.

% A submitted program is expected to be of benefit to other physicists or physical chemists, or be an exemplar of good programming practice, or illustrate new or novel programming techniques which are of importance to some branch of computational physics or physical chemistry.
%
% Acceptable program descriptions can take different forms. The following Long Write-Up structure is a suggested structure but it is not obligatory. Actual structure will depend on the length of the program, the extent to which the algorithms or software have already been described in literature, and the detail provided in the user manual.

%Your manuscript and figure sources should be submitted through the Elsevier Editorial System (EES) by using the online submission tool at \\
% http://www.ees.elsevier.com/cpc.

%In addition to the manuscript you must supply: the program source code; job control scripts, where applicable; a README file giving the names and a brief description of all the files that make up the package and clear instructions on the installation and execution of the program; sample input and output data for at least one comprehensive test run; and, where appropriate, a user manual. These should be sent, via email as a compressed archive file, to the CPC Program Librarian at cpc@qub.ac.uk.

\end{abstract}

%%
%% Start line numbering here if you want
%%
% \linenumbers

% Computer program descriptions should contain the following
% PROGRAM SUMMARY.

{\bf PROGRAM SUMMARY}%/NEW VERSION PROGRAM SUMMARY}
  %Delete as appropriate.

\begin{small}
\noindent
{\em Manuscript Title:}{\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of finite-dimensional and affine Lie algebras                                       \\
{\em Authors:}Anton Nazarov                                                \\
{\em Program Title:}Affine.m                                          \\
{\em Journal Reference:}                                      \\
  %Leave blank, supplied by Elsevier.
{\em Catalogue identifier:}                                   \\
  %Leave blank, supplied by Elsevier.
{\em Licensing provisions:}none                                   \\
  %enter "none" if CPC non-profit use license is sufficient.
{\em Programming language:}Mathematica                                   \\
{\em Computer:}i386-i686, x86\textunderscore 64                                               \\
  %Computer(s) for which program has been designed.
{\em Operating system:} Linux, Windows, MacOS, Solaris                                       \\
  %Operating system(s) for which program has been designed.
{\em RAM:} 5-500 Mb                                              \\
  %RAM in bytes required to execute program with typical data.
%{\em Number of processors used:}                              \\
  %If more than one processor.
%{\em Supplementary material:}                                 \\
  % Fill in if necessary, otherwise leave out.
{\em Keywords:} Mathematica; Lie algebra; affine Lie algebra; Kac-Moody algebra; root system; weights; irreducible modules, CFT, Integrable systems\\
  % Please give some freely chosen keywords that we can use in a
  % cumulative keyword index.
{\em Classification:} 5 Computer Algebra, 4.2 Other algebras and groups                                         \\
  %Classify using CPC Program Library Subject Index, see (
  % http://cpc.cs.qub.ac.uk/subjectIndex/SUBJECT_index.html)
  %e.g. 4.4 Feynman diagrams, 5 Computer Algebra.
%{\em External routines/libraries:}                            \\
  % Fill in if necessary, otherwise leave out.
%{\em Subprograms used:}                                       \\
  %Fill in if necessary, otherwise leave out.
%{\em Catalogue identifier of previous version:}*              \\
  %Only required for a New Version summary, otherwise leave out.
%{\em Journal reference of previous version:}*                  \\
  %Only required for a New Version summary, otherwise leave out.
%{\em Does the new version supersede the previous version?:}*   \\
  %Only required for a New Version summary, otherwise leave out.
{\em Nature of problem:}\\
  %Describe the nature of the problem here.
Representation theory of finite-dimensional Lie algebras has many applications in different branches of physics, including elementary particle physics, molecular physics, nuclear physics. Representations of affine Lie algebras appear in string theories and two-dimensional conformal field theory used for the description of critical phenomena in two-dimensional systems. Also Lie symmetries play major role in study of quantum integrable systems.
   \\
{\em Solution method:}\\
  %Describe the method solution here.
We work with weights and roots of finite-dimensional and affine Lie algebras and use Weyl symmetry extensively. Central problems which are the computations of weight multiplicities, branching and fusion coefficients are solved using one general recurrent algorithm based on generalization of Weyl character formula. We also offer alternative implementation based on Freudenthal multiplicity formula which can be faster in some cases.
   \\
%{\em Reasons for the new version:}*\\
  %Only required for a New Version summary, otherwise leave out.
%   \\
%{\em Summary of revisions:}*\\
  %Only required for a New Version summary, otherwise leave out.
%   \\
{\em Restrictions:}\\
  %Describe any restrictions on the complexity of the problem here.
Computational complexity grows fast  with the rank of algebra, so computations for algebra of rank greater than 8 are non practical.
   \\
{\em Unusual features:}\\
  %Describe any unusual features of the program/problem here.
We offer the possibility to use traditional mathematical notation for objects in representation theory of Lie algebras in computations if {\bf Affine.m} is used in {\it Mathematica} notebook interface.
   \\
%{\em Additional comments:}\\
  %Provide any additional comments here.
%   \\
{\em Running time:}\\
  %Give an indication of the typical running time here.
From seconds to days depending on rank of algebra and complexity of representations.
   \\

% \begin{thebibliography}{0}
% \bibitem{1}Reference 1         % This list should only contain those items referenced in the
% \bibitem{2}Reference 2         % Program Summary section.
% \bibitem{3}Reference 3         % Type references in text as [1], [2], etc.
%                                % This list is different from the bibliography at the end of
%                                % the Long Write-Up.
% \end{thebibliography}
% * Items marked with an asterisk are only required for new versions
% of programs previously published in the CPC Program Library.\\
\end{small}

%% main text
\section{Introduction}
\label{intro}

Representation theory of Lie algebras is of central importance for different areas of physics and mathematics. In physics Lie algebras are used to describe symmetries of quantum and classical systems. Computational methods in representation theory have a long history \cite{belinfante1989survey}, there exist numerous software packages for computations related to Lie algebras \cite{simplie}, \cite{vanleeuwen1994lsp}, \cite{stembridge1995mps,coxweyl}, \cite{fischbacher2002ilp}, \cite{Fuchs:1996dd}.

Most popular programs \cite{simplie}, \cite{vanleeuwen1994lsp}, \cite{fischbacher2002ilp}, \cite{coxweyl} are created to study representation theory of simple finite-dimensional Lie algebras. The main computational problems are the following:
\begin{enumerate}
\item Construction of root system which determine the properties of Lie algebra including its commutation relations.
\item Weyl group traversal which is important due to Weyl symmetry of root system and characters of representations.
\item Calculation of weight multiplicities, branching and fusion coefficients, which are essential for construction and study of representations.
\end{enumerate}
There are well-known algorithms for these tasks \cite{moody1982fast}, \cite{stembridge2001computational}, \cite{belinfante1989survey}, \cite{casselman1994machine}.
The third problem is most computation-intensive. There are two different recurrent algorithms which are based on the Weyl character formula and the Freudenthal multiplicity formula. In this paper we analyze them both.

Infinite-dimensional Lie algebras also have growing number of applications in physics for example in conformal field theory and in a study of  integrable systems. But infinite-dimensional algebras are much harder to investigate and the number of available computer programs is much smaller in this case.

Affine Lie algebras \cite{kac1990idl} constitute important and tractable class of infinite-dimensional Lie algebras. They are constructed as central extensions of loop algebras of (semi-simple) finite-dimensional Lie algebras and appear naturally in a study of Wess-Zumino-Witten and coset models in conformal field theory \cite{Walton:1999xc}, \cite{difrancesco1997cft}, \cite{Goddard198588}, \cite{Dunbar:1992gh}.

The structure of affine Lie algebras allows to adapt for them computational algorithms created for finite-dimensional Lie algebras  \cite{Fuchs:1996dd}, \cite{gannon2001algorithms}, \cite{kass1990ala}. The book \cite{kass1990ala} with the tables of multiplicities and other computed characteristics of affine Lie algebras and representations was published in 1990. But we are not aware of software packages for popular computer algebra systems which can be used to extend these results.
We address this issue and present {\bf Affine.m} -- {\it Mathematica} package for computations in representation theory of affine and finite-dimensional Lie algebras.  We describe the features and limitations of the package in the present paper.  We also provide representation-theoretical background of implemented algorithms and present examples of computations relevant to physical applications.

The paper starts with an overview of Lie algebras and their representation theory (Sec. \ref{sec:theor-backgr}). Then we describe datastructures of {\bf Affine.m} used to present different objects related to Lie algebras and representations (Sec. \ref{sec:core-datastructures}) and discuss implemented algorithms (Sec. \ref{sec:comp-algor}). Next section consists of physically interesting examples (Sec. \ref{sec:examples}). The paper is concluded with the discussion of possible extensions and refinements (Sec. \ref{sec:conclusion}).

\section{Theoretical background}
\label{sec:theor-backgr}

In this section we remind necessary definitions and present formulae used in computations.

\subsection{Lie algebras of finite and affine types}
\label{sec:lie-algebras-finite}

{\it Lie algebra} $\gf$ is a vector space with bilinear operation $[\cdot,\cdot]:\gf\otimes\gf\to \gf$, which is called {\it Lie bracket}. If we choose the basis $X_{i}$ in $\gf$ we can specify commutation relations by the {\it structure constants} $C_{ijk}$:
\begin{equation}
  \label{eq:1}
  [X^{i},X^{j}]=\sum_{k} C^{ij}_{k} X^{k}
\end{equation}
Lie algebra is {\it simple} if it contains no non-trivial ideals with respect to commutator. {\it Semisimple} Lie algebra is a direct sum of simple Lie algebras. In the present paper we treat simple and semisimple Lie algebras.

{\it Cartan subalgebra}  $\hfg$ is a nilpotent subalgebra of  $\gf$ that coincides with its normalizer. 
We denote the elements of basis of $\hfg$ by $H^{i}$.

Killing form on $\gf$ gives a non-degenerate bilinear form $(\cdot,\cdot)$ on $\hfg$ which can be used to identify $\hfg$ with the subspace of the dual space $\hfg^{*}$ of linear functionals on $\hfg$. {\it Weights} are the elements of $\hfg^{*}$ and are denoted by Greek letters $\mu,\nu, \omega, \lambda\dots$

Special choice of basis gives compact description of commutation relations (\ref{eq:1}). This basis can be encoded by the root system which is discussed in Section \ref{sec:weights-roots} (See also \cite{humphreys1997introduction,humphreys1992reflection}).

{\it Loop algebra} $L\gf=\gf\otimes \mathbb{C}[t,t^{-1}]$, corresponding to semisimple Lie algebra $\gf$, has commutation relations
\begin{equation}
  \label{eq:6}
  [X^{i}t^{n},X^{j}t^{m}]=t^{n_+m}\sum_{k}C^{ij}_{k}X^{k}
\end{equation}
Central extension leads to the appearance of an additional term
\begin{equation}
  \label{eq:7}
   [X^{i}t^{n}+\alpha c,X^{j}t^{m}+\beta c]=t^{n+m}\sum_{k}C^{ij}_{k}X^{k}+(X^{i},X^{j})n\delta_{n+m,0}c
\end{equation}
This algebra $\hat\gf=\gf\otimes\mathbb{C}[t,t^{-1}]\oplus\mathbb{C}c$ is called non-twisted {\it affine Lie algebra} \cite{kac1990idl}, \cite{wakimoto2001idl,wakimoto2001lectures}, \cite{kass1990ala}.

We do not treat twisted affine Lie algebras in the present paper.

\subsection{Modules, weights and roots}
\label{sec:weights-roots}

Let $\gf$ be a inite-dimensional or an affine Lie algebra.

Then $\gf$-module is a vector space $V$ together with a bilinear map $\gf \times V\to V$ such that
\begin{equation}
  \label{eq:2}
  [x,y]\cdot v = x\cdot(y\cdot v) - y\cdot(x\cdot v), \quad \mbox{for}\; x,y\in \gf, v\in V
\end{equation}
Representation of algebra $\gf$ on a vector space $V$ is the homomorphism $\gf\to gl(V)$ from $\gf$ to Lie algebra of endomorphisms of the  vector space $V$ with the commutator as a bracket.

For an arbitrary representation it is possible to diagonalize the operators corresponding to Cartan generators $H^{i}$ simultaneously by a special choice of basis $\{v_{j}\}$ in $V$:
\begin{equation}
  \label{eq:3}
  H^{i}\cdot v_{j}=\nu_{j}^{i}v_{j}
\end{equation}
Eigenvalues $\nu^{i}_{j}$ of Cartan generators on an element of basis $v_{j}$ determine a weight $\nu_{j}\in \hfg^{*}$ such that $\nu_{j}(H^{i})=\nu_{j}^{i}$. Vector $v\in V$ is called the weight vector of the weight $\lambda$ if $H v=\lambda_{j}(H)v,\; \forall H\in \hf$. The  weight subspace consists of all weight vectors $V_{\lambda}=\{v\in V: H v=\lambda_{j}(H)v,\; \forall H\in \hf\}$. The weight multiplicity $m_{\lambda}=\mathrm{mult}(\lambda)=\mathrm{dim} V_{\lambda}$ is the dimension of the weight subspace.

The structure of module is determined by the set of weights since the action of generators $E^{\alpha}$ on weight vectors is
\begin{equation}
  \label{eq:5}
  E^{\alpha}\cdot v_{\lambda} \propto v_{\lambda+\alpha}
\end{equation}
Module structure can be encoded by the formal character of a module
\begin{equation}
  \label{eq:10}
  \mathrm{ch}V=\sum_{\lambda}m_{\lambda} e^{\lambda}
\end{equation}
Character  $\mathrm{ch}V\in \mathcal{E}$ is an element of algebra $\mathcal{E}$ generated by formal exponents of weights.
A character can be specialized by taking its value on some element $\xi$ of $\hf$.

Lie algebra is its own module with respect to a special kind of representation. The action that
defines this representation is called adjoint and is given by the bracket $ad_{X} Y=[X,Y]$. 
{\it Roots} are weights of the adjoint representation of $\gf$. They encode the commutation relations of algebra in the following way. Denote by $\Delta$ the set of roots. For each $\alpha\in \Delta$ there exist a root $-\alpha\in \Delta$ and the generators $E^{\alpha}, E^{-\alpha}$ such that
\begin{align}
  \label{eq:4}
  &  [H^{i},E^{\alpha}]=\alpha^{i}E^{\alpha} \\
  &\left[E^{\alpha},E^{\beta}\right]=
  \begin{cases}
    N_{\alpha,\beta} E^{\alpha+\beta}, & \mbox{if}\; \alpha+\beta\in \Delta\\
    \frac{2}{(\alpha,\alpha)} \sum_{i}\alpha^{i} H^{i},&  \mbox{if}\; \alpha=-\beta\\
    0,&\mbox{otherwise}
  \end{cases}
\end{align}

Given the root system $\Delta$ we can choose the set of positive roots. This is a subset  $\Delta^{+}\subset \Delta$ such that for each root $\alpha\in\Delta$ exactly one of the roots $\alpha, -\alpha$ is contained in $\Delta^{+}$ and for any two distinct positive roots $\alpha, \beta\in \Delta^{+}$ such that $\alpha+\beta\in \Delta$ their sum is also positive $\alpha+\beta\in\Delta^{+}$.
Elements of $-\Delta^{+}$ are called negative roots.

A positive root is {\it simple} if it cannot be written as a sum of positive roots. The set of simple roots $\Phi=\left\{\alpha_{i}\right\}$ is a basis in $\hfg^{*}$ and each root can be written as $\alpha=\sum_{i}n_{i}\alpha_{i}$ with all $n_{i}$ non-negative or non-positive simultaneously. In case of a finite-dimensional Lie algebra $\gf$ simple roots are numbered from 1 to the rank of algebra $i=1,\dots,r,\quad r=\mathrm{rank}(\gf)$. By numbering simple roots with an index $i$ we introduce lexicographic ordering in the root system $\Delta$. Highest root with respect to this ordering is denoted by  $\theta=\sum_{i=1,\dots,r} a_i \alpha_i$, coefficients $a_i$ are called {\it marks}. $\theta$ is also the highest weight  of the adjoint module (See section \ref{sec:high-weight-modul}). {\it Comarcs} are the numbers equal to $a_i^{\vee}=\frac{(\alpha_i,\alpha_i)}{2} a_i$.

Although for affine Lie algebra $\hat\gf$ the set of roots $\Delta$ is infinite the set of simple roots $\Phi$ is finite and its elements are denoted by $\alpha_{0},\dots \alpha_{r}$ where $r=\mathrm{rank}(\gf)$. The roots $\alpha_1,\dots, \alpha_r$ are the roots of the underlying finite-dimensional Lie algebra $\go$. The root $\alpha_0=\delta-\theta$ is the difference of {\it imaginary root} $\delta$ and $\theta$ -- the highest root of the algebra $\go$.
Note that root multiplicity $\mathrm{mult}(\alpha)$ for an affine Lie algebra root can be greater than one.
